---
title: "EDE Final Project"
subtitle: "Plastic Additives Adversely Affecting Aquatic Life"
author: "Jalal Bayar & Danielle Butler"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage

```{r setup, include=FALSE}
#Set working directory
getwd()
#Load packages
library(lubridate); library(tidyverse); library(corrplot); library(agricolae);
library(here); library(dplyr)

#Set ggplot theme
mytheme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

#Load dataset
ET.Pht <- read.csv(
  file = here("./ECOTOX_Phthalates_Aquatics_raw.csv"),
  stringsAsFactors = TRUE,
  colClasses = c("CAS.Number" = "factor")
)

```

# 1 Rationale and Research Questions
We wanted to look into the differences in the NOEC concentration of Di-ethyl-
hexyl-phthalate (DEHP), an abundant plasticizer in plastics, across species of
fish. NOEC stands for "No Observed Effect Concentration" which means it is the
highest tested concentration of a substance where no statistically significant
adverse effect is observed. It's used in ecotoxicology studies to determine the 
lowest concentration at which a chemical might be considered safe for a
particular species or ecosystem. 

If we had found out that the NOEC concentration across all the species was
the same, we could have provided a generic threshold limit for all the species.
This analysis would be most helpful from a risk assessment perspective. If we
could not find the same NOEC concentration from multiple studies for various fish
species, then we could not give a generic threshold. Also, the life stage is
important for all the organisms. We are looking to see if the NOEC concentration
of DEHP is the same for embryonic, larval, and adult life stage or not.

We want to look to answer specific questions such as:
-Which fish species is most sensitive to DEHP?
-Which life stage of the fish species is most sensitive to DEHP?
-For our Anova analysis, our Null Hypothesis is: there is no significant
difference between the fish species and their life stage.
-And the Alternate Hypothesis: there will be a significant difference between
the fish species and their life stage. 

```{r,include=FALSE}
#This code chunk contains wrangling of fish, crustaceans, and amphibians.
#Wrangling fish
fish <- ET.Pht %>% 
  filter(Chemical.abbreviation %in% c("DEHP")) %>% 
  filter(Species.Group %in% c("Fish", "Fish; Standard Test Species",
                              "Fish; Standard Test Species; U.S. Threatened and
                              Endangered Species"
                              )) %>% 
  filter(Organism.Lifestage %in% c("Embryo", "Larva", "Adult")) %>% 
  filter(Endpoint %in% c("NOEC")) %>% 
  filter(Conc.1.Units..Standardized. %in% c("AI mg/L",
                                            "AI mg/kg bdwt"))

#Select appropriate columns
fish_tidy <- fish %>% 
  select("Chemical.abbreviation", 
         "Species.Common.Name",
         "Organism.Lifestage",
         "Endpoint",
         "Conc.1.Mean..Standardized."
         )

#Rename columns
colnames(fish_tidy) <- c('phthalate', 'species', 'life_stage', 'endpoint',
                              'concentration')

#Wrangle crustaceans
crustaceans <- ET.Pht %>% 
  filter(Chemical.abbreviation %in% c("DEHP")) %>% 
  filter(Species.Group %in% c("Crustaceans",
                              "Crustaceans; Standard Test Species"
                              )) %>% 
  filter(Organism.Lifestage %in% c("Juvenile", "Adult")) %>% 
  filter(Endpoint %in% c("NOEC")) %>% 
  filter(Conc.1.Units..Standardized. %in% c("AI mg/L",
                                            "AI mg/kg bdwt"))

#Select appropriate columns
crustaceans_tidy <- crustaceans %>% 
  select("Chemical.abbreviation", 
         "Species.Common.Name",
         "Organism.Lifestage",
         "Endpoint",
         "Conc.1.Mean..Standardized."
         )

#Rename columns
colnames(crustaceans_tidy) <- c('phthalate', 'species', 'life_stage', 'endpoint'
                                ,'concentration')

#Amphibians wrangling
amphibians <- ET.Pht %>% 
  filter(Chemical.abbreviation %in% c("DEHP")) %>% 
  filter(Species.Group %in% c("Amphibians",
                              "Amphibians; Standard Test Species"
                              )) %>% 
  filter(Organism.Lifestage %in% c("Embryo", "Blastula")) %>% 
  filter(Endpoint %in% c("NOEC")) %>% 
  filter(Conc.1.Units..Standardized. %in% c("AI mg/L",
                                            "AI mg/kg bdwt"))

#Select appropriate columns
amphibians_tidy <- amphibians %>% 
  select("Chemical.abbreviation", 
         "Species.Common.Name",
         "Organism.Lifestage",
         "Endpoint",
         "Conc.1.Mean..Standardized."
         )

#Rename columns
colnames(amphibians_tidy) <- c('phthalate', 'species', 'life_stage', 'endpoint',
                              'concentration')
```

\newpage

# 2 Dataset Information
We extracted data from the eco-toxicolgy knowledge-base resource hub. This
dataset is regulated by the U.S. EPA and contains over 13,000 chemicals, and
14,000 species. We chose phthalate esters as the chemical to inspect the
concentrations of. We downloaded data for phthalates for 16 classes of species,
such as fish, crustaceans, amphibians, insects/spiders and more. See the link
here for more details: https://cfpub.epa.gov/ecotox/. We wrangled to look at
Fish, Amphibians and Crustaceans, and then decided to deep dive into Fish. We
also filtered to only look at DEHP at the NOEC levels of concentration.

\newpage

# 3 Exploratory Analysis 
```{r}
#Raw data dimensions and column names
dim(ET.Pht)
colnames(ET.Pht)

#We are interested only in DEHP, fish, NOEC, and life stages. 
#Selected the following columns: Chemical.abbreviation, Species.Common.Name, 
#Organism.Lifestage, Endpoint, and Conc.1.Mean..Standardized.

```
# 4 Visualizations

```{r}
#Data Exploration and Visualization 

#Fish Boxplot
ggplot(fish_tidy, aes(x = species, y = concentration, color = life_stage)) +
  geom_boxplot() +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Fish - DEHP NOEC Concentrations") +
  labs(x = "Species",
       y = expression(Log[10] * NOEC * " [mg/L]"), 
       color = "Life stage")

#Amphibians Boxplot
ggplot(amphibians_tidy, aes(x = species, y = concentration, color = life_stage))+
  geom_boxplot() +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Amphibians - DEHP NOEC Concentrations") +
  labs(x = "Species",
       y = expression(Log[10] * NOEC * " [mg/L]"), 
       color = "Life stage")

#Crustaceans Boxplot
ggplot(crustaceans_tidy, aes(x = species, y = concentration, color = 
                               life_stage)) +
  geom_boxplot() +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Crustaceans - DEHP NOEC Concentrations") +
  labs(x = "Species",
       y = expression(Log[10] * NOEC * " [mg/L]"), 
       color = "Life stage")

```

\newpage

# 5 Analysis
Ho: The mean NOEC concentration is the same across the fish species and their
life stages.

Ha: There is a significant difference between the fish species and their
life stages.

```{r}
#ANOVA DEHP
anova <- aov(data = fish_tidy, concentration ~ life_stage + species)
summary(anova)

#These anova results show that life stage and species are both significantly
#different with life stage having a stronger p value than species.

#HSD DEHP Species
hsd.test.species <- HSD.test(anova, "species",
group = TRUE)
hsd.test.species

#The output of the HSD test for species specifically shows that the Zebra Danio
#was in its own group compared to the rest of the species.

#HSD DEHP Life Stage
hsd.test.lifestage <- HSD.test(anova, "life_stage",
group = TRUE)
hsd.test.lifestage

#The output of the HSD test for life stage specifically shows that the Embryo 
#was in its own group compared with the Larva and Adult life stages.

#For two-way anova
TukeyHSD(anova)

```

\newpage

# 6 Summary and Conclusions
Questions:
-Which fish species is most sensitive to DEHP?
-Which life stage of the fish species is most sensitive to DEHP?

We found a significant difference between the fish species and their life
stages, which means, the NOEC concentrations vary within these species. Also,
they vary between their life stages. So, we could consider the lowest NOEC
concentration that we found for Guppy for all fish species, but that would be
a very risk averse option for some of the larger fish.

For future studies, life stage is an important parameter to consider for all the
chemicals and all the species. It would be interesting to bring in other 
variables that might affect the NOEC concentration level in the future like
weight or length of the fish species, for example.

\newpage

# 7 References
<Olker, J. H., Elonen, C. M., Pilli, A., Anderson, A., Kinziger, B., Erickson, S.,
Skopinski, M., Pomplun, A., LaLone, C. A., Russom, C. L., & Hoff, D. (2022). The
ECOTOXicology Knowledgebase: A Curated Database of Ecologically Relevant
Toxicity Tests to Support Environmental Research and Risk
Assessment. Environmental Toxicology and Chemistry, 41(6):1520-1539.
https://doi.org/10.1002/etc.5324> 
